### 2025-05-09 Chapters/Chapter3.tex
vinna

\chapter{Results}
\label{Chapter4}
\section{Baseline models}
The simplest model for the gust factor is to guess the average gust factor for all available observations, i.e. use no variables from the training data. A slightly better model is obtained by using the wind speed as a training feature, either through regression or NN training. The MAPEs for the constructed base models are shown in Table \ref{table:baseline_models}. The ws$_{15}$ model was obtained with NN training but regression produces an almost identical result.

@@

\section{Modeling for various ws$_\textbf{15}$ lower bounds}
To assess how model performance depends on ws$_{15}$ lower bound several models were trained using the primary variable set with and without DEM. The results can be seen in Table \ref{table:results}. For this investigation comparison was only made with the constant baseline model. In all cases adding the DEM information is seen to increase the model performance. The reason for the difference between the 10 m/s lower bound results seen in Tables \ref{table:setsOfParams} and \ref{table:results} is they were obtained with different training setups.

\begin{table}[h]
  \caption[Model results for different wind speed limits]{MAPE for each ws$_{15}$ lower bound without and with landscape elevation data (DEM) in a 30° sector into the direction of the reanalysis wind. Adding DEM data reduces the error noticably. Note that the MAPE is higher for both the lower lower bounds and for the higher ones.}
    \label{table:results}
    \centering
    \begin{tabular}{lccc}
        \toprule
        Wind Speed & \multicolumn{3}{c}{MAPE} \\
        limit $[m/s]$ & Baseline & Without DEM & With DEM \\
        \midrule
        $\geq 0$ & 39.2\% & 19.2\% & 18.9\% \\
        $\geq 5$ & 28.1\% & 15.3\% & 14.9\%\\
        $\geq 10$ & 23.9\% & 13.3\% & 12.5\%\\
        $\geq 15$ & 23.2\% & 14.4\% & 11.9\%\\
        $\geq 20$ & 24.7\% & 15.7\% & 13.3\%\\
        $\geq 25$ & 27.7\% & 19.4\% & 17.3\%\\
        \bottomrule
    \end{tabular}
  \end{table}

@@

>>>>

\chapter{Results}
\label{Chapter3}
\section{Baseline models}
The simplest model for the gust factor is to guess the average gust factor for all available observations, i.e. use no variables from the training data. A slightly better model is obtained by using the wind speed as a training feature, either through regression or NN training. The MAPEs for the constructed base models are shown in Table \ref{table:baseline_models}. The ws$_{15}$ model was obtained with NN training but regression produces an almost identical result.

@@

\section{Modeling for various ws$_\textbf{15}$ lower bounds}
To assess how model performance depends on ws$_{15}$ lower bound, several models were trained using the primary variable set, with and without DEM. The results can be seen in Table \ref{table:results}. For this investigation comparison was only made with the constant baseline model. In all cases adding the DEM information is seen to increase the model performance. The reason for the difference between the 10 m/s lower bound results seen in Tables \ref{table:setsOfParams} and \ref{table:results} is that they were obtained with different training setups.

\begin{table}[h]
  \caption[Model results for different wind speed limits]{MAPE for each ws$_{15}$ lower bound without and with landscape elevation data (DEM) in a 30° sector into the direction of the reanalysis wind. Adding DEM data reduces the error noticably. Note that the MAPE is worse for both the lower lower bounds and the higher ones, but better for the intermediate ones.}
    \label{table:results}
    \centering
    \begin{tabular}{lccc}
        \toprule
        Wind Speed & \multicolumn{3}{c}{MAPE} \\
        limit $[m/s]$ & Baseline & Without DEM & With DEM \\
        \midrule
        $\geq 0$ & 39.2\% & 19.2\% & 18.9\% \\
        $\geq 5$ & 28.1\% & 15.3\% & 14.9\%\\
        $\geq 10$ & 23.9\% & 13.3\% & 12.5\%\\
        $\geq 15$ & 23.2\% & 14.4\% & 11.9\%\\
        $\geq 20$ & 24.7\% & 15.7\% & 13.3\%\\
        $\geq 25$ & 27.7\% & 19.4\% & 17.3\%\\
        \bottomrule
    \end{tabular}
  \end{table}

@@

\section{CARRA wind gust}
% In 4.3, add commas in first sentence
% Later in 4.3, add "that" before "they"
% Table 4.3, change higher to worse, add: but better for the intermediate ones

The CARRA dataset provides no gust values in its analysis component but does include 10 m wind gust in the forecast. With a shortest lead time of one hour and given that a model of the gust factor is to be constructed, it stands to reason to incorporate the one-hour forecast CARRA gust factor, computed as:
\begin{equation*}
\text{CARRA gust factor} = \dfrac
{\text{1 hr forecast 10 m wind gust}}
{\text{1 hr forecast 10 m wind speed}}
\end{equation*}
into the modelling, as a proxy for a reanalysis wind gust.

One issue with the CARRA gust factor is that it is not available for several time points including the period after the middle of 2022, whereas the CARRA reanalysis data used in this study extends through 2023, making it somewhat involved to assess its predictive power. Two investigations were carried out. First, the reduction in MAPE resulting from the addition of the CARRA gust factor to the baseline ws$_{15}$, secondary, and primary NN models of Section 4.2 was assessed, using only data with CARRA gust available. The results of this comparison are shown in Table \ref{table:carra_gust}

\begin{table}[h]
  \centering
  \caption[Results adding CARRA gust factor]{Results of adding the CARRA gust factor as a variable in selected models from Sections 4.2 and 4.3.}
  \label{table:carra_gust}
  \begin{tabular}{lcc}
    \toprule
    & \multicolumn{2}{c}{MAPE\qquad\qquad} \\
    Model & Unaltered & With CARRA gust factor \\
    \midrule
    Baseline ws$_{15}$ model & 17.0\% & 16.7\% \\
    Secondary model           & 13.3\% & 13.2\% \\
    Primary model             & 12.5\% & 12.4\% \\
    \bottomrule
  \end{tabular}
\end{table}

The causes of the differences between the numbers in Table \ref{table:carra_gust} and those of Tables 4.2 and 4.3 could be different time period and/or unlike training setup.

Second, three simple linear regression models were constructed involving ws$_{15}$ and the CARRA gust factor, using available data with ws$_{15} \geq 10$ m/s. The results are shown in Table \ref{table:carra_gust_regression}.

\begin{table}[h]
  \centering
  \caption[Regression MAPE with and without CARRA gust factor]{Mean absolute percentage error (MAPE) for regression models using the CARRA gust factor, ws$_{15}$, and both combined.}
  \label{table:carra_gust_regression}
  \begin{tabular}{lc}
    \toprule
    Model                         & MAPE   \\
    \midrule
    ws$_{15}$                     & 17.03\% \\
    CARRA–gf                      & 16.81\% \\
    CARRA–gf + ws$_{15}$          & 16.80\% \\
    \bottomrule
  \end{tabular}
\end{table}

Finally, using the raw CARRA gust factor without any modelling gave MAPE = 18.14\%.

The conclusion that can be drawn from this investigation is that there is a modest information in the CARRA wind gust forecast, potentiially reducing model MAPE by 0.1 to 0.3 percentage points. This reduction is however quite small compared with the overall MAPE reduction of $\sim$5 percentage points seen in Table 4.2 when all available variables are added.

### 2025-05-09 MSc.tex
búnir með kafla 4
Updated and added neural network figure.
vinna

\def\thesisyear{2025}       					% Year thesis submitted
\def\thesismonth{June}					% Month thesis submitted, e.g. "December"
\def\thesisauthor{Brynjar Geir Sigurðsson}				% Thesis authoreiningaraðferðinni
\def\thesistitle{A Neural Network Approach to Predicting Gust Factors in Complex Landscape} % Title of thesis
%\def\thesissubtitle{XXSubtitle is rarely usedXX}		% Subtitle of thesis (optional)
\def\thesisshorttitle{} 	% Optional: if title of thesis is longer than 50 characters, it would not fit on the spine (kjölur) and in this case you, need to provide a short title here for the print shop. Otherwhise: make it empty
\def\thesiscredits{60} 						% Credits awarded for the project
\def\thesissubject{Mechanical Engineering}
\def\thesiskind{M.Sc.}					% E.g. M.Sc. or Ph.D. thesis
\def\thesiskindformal{Magister Scientiarum}			% E.g. Magister Scientiarum
\def\thesisschool{School of Engineering and {Natural Sciences}}		% School
\def\thesisfaculty{Industrial Engineering, Mechanical Engineering and Computer Science}% Faculty name without "Faculty of" (gets added) 
\def\thesisaddress{Dunhagi 5}			        % Faculty or school office address street
\def\thesispostalcode{107}			                % Faculty or school office zip code
\def\thesistelephone{525 4000}					% Office telephone
%\def\thesispublisher{XX}					% Publisher (not used)
\def\thesissupervisors{Kristján Jónasson}					% Names of supervisors (split by \\ if more than one).
\def\thesisnrofsupervisors{1}					% Number of supervisors (to use "Supervisor" vs. "Supervisors")
\def\thesiscommittee{Kristján Jónasson \\ Ólafur Pétur Pálsson \\ Guðrún Nína Petersen}			% Thesis commitee must include the supervisor(s)
\def\thesisexaminer{Halldór Björnsson}				        % Examiner (prófdómari)
\def\thesisISBN{}           					% Thesis ISBN number (keep empty: not used anymore)
\def\thesisprinting{}						% Name of printsthop (keep empty if thesis is never printed)
%\def\thesisprinting{Háskólaprent, Fálkagata 2, 107 Reykjavík}	% Name of printsthop (keep empty if thesis is never printed)
\def\thesislicense{This thesis may not be copied in any form without author permission.} % Set license here (could also be some Creative Commons license)
%\def\thesiskeywords{Keyword1, Keyword2, Keyword3}		% Keywords (not used anywhere, hence commented out)

@@

\section*{Abstract}
In many industries, whether it be windmill farming or transportation, being able to precisely predict the load caused by weather is crucial to prevent harm coming to people and machinery as well as increasing productivity. The efficiency of wind turbines in high wind speed conditions is a vital challenge in optimizing renewable energy systems. Gust predictions can be instrumental in loss prevention. If looking only at wind, the most destructive moments are gusts, by definition. To predict these values there are two ways, traditional numerical weather prediction systems and machine learning (or other pattern recognition) methods. Computer-aided numerical weather prediction systems originated in the 1950s, while the use of machine learning in numerical weather predictions is much more recent, having gained traction in the last decade. The goal of this thesis is to combine these two methods to predict the wind gust factor, defined as the ratio between wind gust and average wind speed. This is done by having the input data to the machine learning model be reanalysis variables that were generated by traditional numerical weather prediction systems. In addition to these variables (and derived variables), a digital elevation model (DEM) was used. A deep neural network was constructed using some set of variables and the results are compared to a baseline as well as each other so as to show the impact each feature has. Neural networks can help make wind gust predictions in complex ladscape.
\vfill \vspace*{1cm}

\section*{Útdráttur}
Hvort sem um er að ræða vindmyllubúgarða eða stöðuleiki flutningabíla er spágetan um álag vegna vinds mikilvæg til að koma í veg fyrir tjón á iðnviðum, auk þess er hægt að auka framleiðni. Hviðuspár geta verið stór þáttur í tjónaminnkun. Ef einungis er horft á vind, þá eru hviður hæsti álagspunktur samkvæmt skilgreiningu. Sögulega hefur verið notast við hefðbundin spálíkön sem reiða sig á töluleg eðlisfræðileg líkön til að spá fyrir um niðurstöður. Á síðustu árum hefur gervigreind þróast mikið og líkön verið þjálfuð til þess að spá fyrir um veður. Svo miklar framfarir hafa orðið að sum líkön geta keppt við hefðbundin veðurlíkön. Markmið þessa verkefnis er reyna að nýta bæði hefðbundin veðurlíkön og gervigreind til þess að giska á hviðustuðla, þar sem hviðustuðullinn er skilgreindur sem hlutfall hæstu hviðu og meðalvinds. Þetta er gert með því að nota endurgreiningargögn úr hefðbundnu veðurlíkani sem grunngögn í gervigreindarlíkan. Einnig eru notað hæðarlíkan til að lýsa umhverfi. Djúpt tauganet var búið til með því að velja breytur úr endurgreiningargögnunum ásamt hæðarpunktum og afleiddum breytum. Þetta líkan er svo þjálfað og niðurstöður bornar saman við grunnlíkan sem og önnur líkön með öðrum breytum til að skoða áhrif breyta. Djúp tauganet geta búið til hjálplegar spár fyrir vindhviður í flóknu landslagi.
\vfill

@@

\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}
Special thanks go to my advisor, Kristján Jónasson. He would help me with the actual work of the thesis and sit with go over the code with me. Thanks also go to Ólafur Pétur Pálsson, who would read over the thesis with notes and give helpful suggestions. Guðrún Nína Petersen, also provided valuable insight whenever questioned about meteoroligcal matters as well as giving notes on the thesis.

@@

\include{Chapters/Chapter1}
\include{Chapters/Chapter2}
\include{Chapters/Chapter3}
\include{Chapters/Chapter4}
\include{Chapters/Chapter5}

>>>>

\def\thesisyear{2025}       					% Year thesis submitted
\def\thesismonth{June}					% Month thesis submitted, e.g. "December"
\def\thesisauthor{Brynjar Geir Sigurðsson}				% Thesis authoreiningaraðferðinni
\def\thesistitle{Prediction of Gust Factor in Complex Landscape with Neural Networks} % Title of thesis
%\def\thesissubtitle{XXSubtitle is rarely usedXX}		% Subtitle of thesis (optional)
\def\thesisshorttitle{} 	% Optional: if title of thesis is longer than 50 characters, it would not fit on the spine (kjölur) and in this case you, need to provide a short title here for the print shop. Otherwhise: make it empty
\def\thesiscredits{60} 						% Credits awarded for the project
\def\thesissubject{Mechanical Engineering}
\def\thesiskind{M.Sc.}					% E.g. M.Sc. or Ph.D. thesis
\def\thesiskindformal{Magister Scientiarum}			% E.g. Magister Scientiarum
\def\thesisschool{School of Engineering and {Natural Sciences}}		% School
\def\thesisfaculty{Industrial Engineering, Mechanical Engineering and Computer Science}% Faculty name without "Faculty of" (gets added) 
\def\thesisaddress{Dunhagi 5}			        % Faculty or school office address street
\def\thesispostalcode{107}			                % Faculty or school office zip code
\def\thesistelephone{525 4000}					% Office telephone
%\def\thesispublisher{XX}					% Publisher (not used)
\def\thesissupervisors{Kristján Jónasson}					% Names of supervisors (split by \\ if more than one).
\def\thesisnrofsupervisors{1}					% Number of supervisors (to use "Supervisor" vs. "Supervisors")
\def\thesiscommittee{Kristján Jónasson \\ Ólafur Pétur Pálsson \\ Guðrún Nína Petersen}			% Thesis commitee must include the supervisor(s)
\def\thesisexaminer{Halldór Björnsson}				        % Examiner (prófdómari)
\def\thesisISBN{}           					% Thesis ISBN number (keep empty: not used anymore)
\def\thesisprinting{}						% Name of printsthop (keep empty if thesis is never printed)
%\def\thesisprinting{Háskólaprent, Fálkagata 2, 107 Reykjavík}	% Name of printsthop (keep empty if thesis is never printed)
\def\thesislicense{This thesis may not be copied in any form without author permission.} % Set license here (could also be some Creative Commons license)
%\def\thesiskeywords{Keyword1, Keyword2, Keyword3}		% Keywords (not used anywhere, hence commented out)

@@

\section*{Abstract}
Whether it is wind farms, traffic accidents, high-voltage towers, or other things affected by wind, accurately predicting wind loads is crucial for reducing damage, improving operational safety, and boosting productivity. Wind gusts are multi-second spikes in speed, and wind gust factor is defined as the ratio of the highest three-second gust to the ten-minute mean wind speed. It is important to be able to predict both wind and gust factor accurately. Historically, numerical weather prediction models based on physical equations have been used to predict such variables. In recent years, artificial intelligence has advanced rapidly, and AI-based models have begun to rival conventional model performance. This project combines both approaches by using atmospheric model reanalysis output and digital elevation data as inputs to a deep neural network. Selected features from the reanalysis fields, terrain elevation, and derived variables are fed into several network architectures, whose performance is then compared against a simple linear regression baseline. The results are very promising: this hybrid methodology can substantially improve both instantaneous estimation and short-term forecasting of gust factors, particularly in areas of complex terrain.
\vfill \vspace*{1cm}

\section*{Útdráttur}
Hvort sem um er að ræða vindmyllugarða, umferðarslys, háspennumöstur, eða aðra hluti sem vindur getur haft áhrif á er spágeta um álag vegna vindsins mikilvæg, til að minnka eða koma í veg fyrir tjón, auka rekstraröryggi, eða bæta framleiðni. Vindhviður eru nokkurra sekúndna aukning á vindi og hviðustuðull mælir hæsta hlutfall milli þriggja sekúndna hviðu og 10 mínútna meðalvinds. Góðar hviðuspár eru mikilvægar. Sögulega hefur verið notast við hefðbundin spálíkön sem reiða sig á töluleg eðlisfræðileg líkön til að spá fyrir um veður, þar með talið hviðustuðul. Á síðustu árum hefur gervigreind þróast mikið og líkön verið þjálfuð til þess að spá fyrir um veður. Svo miklar framfarir hafa orðið að sum líkön geta keppt við hefðbundin veðurlíkön. Markmið þessa verkefnis er reyna að nýta bæði hefðbundin veðurlíkön og gervigreind til þess að meta hviðustuðla. Þetta er gert með því að nota endurgreiningargögn úr hefðbundnu veðurlíkani sem grunngögn í gervigreindarlíkan. Einnig eru notað hæðarlíkan sem lýsir landslagi. Djúpt tauganet er búið til með því að velja breytur úr endurgreiningargögnunum og hæðarlíkaninu og afleiddum breytum. Nokkur líkön af þessu tagi eru svo þjálfuð og niðurstöður bornar saman við einfalt grunnlíkan sem fengið er með línulega aðhvarfi. Niðurstöðurnar lofa mjög góðu um að þessi aðferðafræði geti bætt bæði samtímamat á hviðustuðli og spár um hann umtalsvert. Þetta á ekki síst við í flóknu landslagi.
\vfill

@@

\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}
Special thanks go to my advisor, Kristján Jónasson, who helped me with the thesis work and sat with me to go over the code. Thanks also to Ólafur Pétur Pálsson who provided detailed notes and offered very helpful suggestions. I am also grateful to Guðrún Nína Petersen, who provided valuable insight into meteorological matters and gave constructive feedback on the thesis.

@@

\include{Chapters/Chapter1}
\include{Chapters/Chapter2}
\include{Chapters/Chapter3}
\include{Chapters/Chapter4}
